Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems

### 2025-04-17 paper/manuscript.Rmd
Last few edits + adding Acknowledgments and data/code repo

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Willem Bonnaffé (d), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (e)"
date: "Research report"
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

Biodiversity temporal changes are expected to profoundly impact ecosystem
functioning and stability, yet large-scale empirical evidence remains limited.
Bridging this gap requires aligning biodiversity temporal trends with food web
theory and Biodiversity-Ecosystem Functioning research, which provide strong
predictions about the ecological consequences of biodiversity loss. Most
insights rely on spatial comparisons of ecological communities, assuming they
reflect temporal trends—a space-for-time substitution. Here, we analyze
biodiversity and food web time series from over 400 riverine fish communities
across France (1995–2018) to evaluate how biodiversity temporal trends reshape
food web structure and community functioning. We examine relationships among
species richness, community biomass, and three key food web metrics:
connectance, weighted average trophic level, and trophic pathway redundancy. We
also test the space-for-time hypothesis using spatial gradients and a
theoretical food web model. Structural equation models and robustness analyses
reveal that declines in species richness strongly correlate with reduced
community biomass, with top trophic-level species playing a disproportionate
role in food web changes. Declining biomass is associated with the loss of top
trophic levels and decreased connectance, while temporal changes of species
richness negatively correlates with those of connectance. Both biomass and
species richness declines reduce trophic pathway redundancy, suggesting
biodiversity loss undermines community robustness to future perturbations. Our
findings align with food web theory and Biodiversity-Ecosystem Functioning
research, support the validity of space-for-time approaches for basic food web
metrics, and highlight the relevance of food web metrics as indicators of
ecosystem function and fragility.

@@

Current rates of biodiversity loss raise urgent concerns about ecosystem
functioning and stability, yet how biodiversity trends directly impact these
critical processes remains unclear. For instance, food web &
Biodiversity-Ecosystem Functioning research delivered strong predictions
about the ecological consequences of biodiversity loss. Using an extensive
dataset of riverine fish communities, we link biodiversity trends to temporal
changes in food web structure. We demonstrate that temporal changes in species
richness and community biomass consistently associate with altered trophic
levels, connectance, and the redundancy of feeding relationships. We found that
declining biodiversity reduces trophic pathway redundancy, likely increasing
ecosystem vulnerability to perturbations. Our findings suggest that food web
structure metrics could serve as valuable indicators for monitoring ecosystem
health and guiding conservation strategies.

@@

(ref:fig-met-cap) Methodological Approach: (a) We analyzed `r nb_st` stream fish
communities in France, each monitored for at least 10 years. (b) For each site
and sampling event, we reconstructed food webs using a metaweb based on
ontogenetic diet information and predator-prey body size ratios. (c) We computed
temporal trends and median values for community biomass, species richness,
connectance, average trophic level, and redundancy at each site. (d) We then
assessed relationships among these food web descriptors over time and (e) across
spatial gradients.

```{r matmet, fig.cap="(fig-met-cap)", out.width="100%"}
knitr::include_graphics(get_mypath("figures", "mat_met.png"))
```

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species richness,
biomass, connectance, weighted average trophic level, and proportion of
redundant trophic links (hereafter “redundancy”, see Appendix 1,
Fig. S1, Table S2) from the `r nb_st`
sampling sites with 10+ years of community dynamics (Fig. \@ref(fig:matmet)a-c). We used
ontogenic food diet data at the species level and individual fish body size to
reconstruct trophic species food webs, i.e. at the size class species level (See
Methods). We implemented Bayesian hierarchical models accounting for spatial
variation at both hydrographic river basin and site levels to jointly estimate
the temporal trends in these metrics (see Methods). We found a great
heterogeneity in the temporal trends at the site level (community biomass: from
`r grbmn`% to +`r grbmx`% per decade, species richness: `r grrhn` to
+`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig.
S2) which was effectively captured by the hierarchical model (conditional R2
ranging from `r r2cond$log_bm_std` for community biomass to
`r r2cond$log_rich_std` for species richness, Table S3).

(ref:sem-cap) (a) Structural equation models linking the temporal trends of
biodiversity and food web structure. Dashed and solid arrows display
respectively negative and positive the standardized slope coefficients, their
widths being proportional to the absolute values of the coefficients. The
standardized slope coefficients and their associated 95% confidence intervals
obtained by bootstrapping are displayed along the path. $R^2$ displays the marginal
R-squared of the linear models. (b to h) Direct bivariate relationships derived
from the structural equation model. Points display the temporal trends at each
site estimated with a hierarchical Bayesian model, lines display the linear
relationship estimated with the SEM. The distribution of the site temporal
trends are displayed in Fig. S2. All the effects are reported in Table S4.
Average trophic level is weighted by the biomass of the trophic species.
Redundancy is the proportion of redundant links.

@@

(ref:fig-std-cap) a, Direct and total effects of community biomass and species
richness trends on food web structure trends. Standardized coefficients (r)
derived from the temporal, spatial, and bioenergetic model Structural Equation
Models. Points and lines display the estimated effects and their confidence
intervals at 95%. Values are displayed in Table S4, R-squared in Table S4, and
the visualization of the bivariate relationships for the spatial and
bioenergetic model SEMs are displayed in Figure S4. Average trophic level is
weighted by the biomass of the trophic species. Redundancy is the proportion of
redundant links. b, Effect of species extinctions on the structure of the
empirical metaweb according to three scenarios. In the Decreasing and Increasing
scenarios, species were ordered by their trophic level and removed from the
metaweb starting respectively from the highest and lowest trophic level. Species
were removed in a random sequence (50 sequences, grey dots) for the random
scenario, the black line displaying the average predicted by a GAM linear model.
For the extinction sequence, the average trophic level was not weighted by the
biomass of species as this information is not available at the metaweb scale.
Similar results were obtained with trophic species (i.e. node) extinction (Fig.
S5).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were standardized and the
confidence intervals were computed with semEff R package [@murphy_semeff_2021]. Confidence
intervals were obtained by bootstrapping the linear models 1000 times and
computing adjusted bootstrap percentile intervals (BCa), which are adjusted for
the asymmetric distribution of resampling. The marginal and conditional R2
associated with the linear models of the SEMs were computed with the Nagelkerke
method [@lefcheck_piecewisesem_2016, Table S5].

@@

We used a bioenergetic food web model to simulate the relations between
community biomass, species richness and food web structure. This food web model
lifts one of the limitations of our data, that is it includes the biomass of all
species in the food webs from primary producers to top predators. This model
allows simulating complex communities because its parameterization depends
mainly on species body masses. We first generated `r nfw` food webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food web topology typically found in empirical
food webs [@williams_simple_2000]. We set species body masses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and body mass empirically estimated
[@brose_allometric_2006], such as $x_i = 0.88 \times M_i^{-0.25}$.

@@

>>>>

---
title: "Biodiversity temporal trends are reshaping food web structure and redundancy in riverine ecosystems"
subtitle: "Alain Danet (a,b), Loubna El Madouri (a,c), Willem Bonnaffé (d), Andrew P. Beckerman (b),
Maud Mouchet (a), Colin Fontaine (a) and Elisa Thébault (e)"
date: "Research report"
output:
  bookdown::pdf_document2:
    fig_caption: true
    keep_tex: true
    toc: false
  officedown::rdocx_document:
    number_sections: false
    base_format: "bookdown::word_document2"
always_allow_html: true
number_sections: false
bibliography: references.bib
link-citations: true
#https://pkgs.rstudio.com/rticles/articles/examples.html
toc: false
linestretch: 2.0
header-includes:
  - \usepackage{amsmath}
  - \usepackage[left]{lineno}
  - \linenumbers
  - \usepackage{hyperref}
  - \hypersetup{backref=true}
  - \usepackage{setspace}\doublespacing
csl: pnas.csl
---

@@

Biodiversity temporal changes are expected to profoundly impact ecosystem
functioning and stability, yet large-scale empirical evidence remains limited.
Bridging this gap requires aligning biodiversity temporal trends with food web
theory and Biodiversity-Ecosystem Functioning research, both of which provide
strong predictions about the ecological consequences of biodiversity loss. Most
insights on those consequences currently rely on spatial comparisons of
ecological communities, assuming they reflect temporal trends—a space-for-time
substitution. Here, we analyze biodiversity and food web time series from over
400 riverine fish communities across France (1995–2018) to evaluate how
biodiversity temporal trends reshape food web structure and community
functioning. We examine relationships among species richness, community biomass,
and three key food web metrics: connectance, weighted average trophic level, and
trophic pathway redundancy. We also test the space-for-time hypothesis using
spatial gradients and a theoretical food web model. Our structural equation
models and robustness analysis reveal that declines in species richness strongly
correlate with reduced community biomass, with top trophic-level species playing
a disproportionate role in food web changes. Declining biomass is associated
with the loss of top trophic levels and decreased connectance, while temporal
changes in species richness negatively correlates with changes in connectance.
Both biomass and species richness declines reduce trophic pathway redundancy,
suggesting that biodiversity loss weakens community robustness to future
perturbations. Our findings align with food web theory and
Biodiversity-Ecosystem Functioning research, support the validity of
space-for-time approaches for basic food web metrics, and demonstrate the
relevance of food web metrics as indicators of ecosystem function and fragility.

@@

Rates of biodiversity loss raise urgent concerns about ecosystem functioning and
stability, yet how biodiversity trends directly impact these critical processes
remains unclear. For instance, food web & Biodiversity-Ecosystem Functioning
research has delivered strong predictions about the ecological consequences of
biodiversity loss but lack of empirical assessment. Using an extensive dataset
of riverine fish communities, we link biodiversity trends to temporal changes in
food web structure. We demonstrate that temporal changes in species richness and
community biomass consistently associate with altered food web structure. We
found that declining biodiversity reduces trophic pathway redundancy, likely
increasing ecosystem vulnerability to perturbations. Our findings suggest that
food web structure metrics could serve as valuable indicators for monitoring
ecosystem health and guiding conservation strategies.

@@

(ref:met-cap) **Methodological approach**: (a) We analyzed `r nb_st` stream fish
communities in France, each monitored for at least 10 years. (b) For each site
and sampling event, we reconstructed food webs using a metaweb based on
ontogenetic diet information and predator-prey body size ratios. (c) We computed
temporal trends and median values for community biomass, species richness,
connectance, average trophic level, and redundancy at each site. (d) We then
assessed relationships among these food web descriptors over time and (e) across
spatial gradients.

```{r matmet, fig.cap="(ref:met-cap)", out.width="100%"}
knitr::include_graphics(get_mypath("figures", "mat_met.png"))
```

@@

Our food web reconstruction delivered `r nb_op` annual estimates of species richness,
biomass, connectance, weighted average trophic level, and proportion of
redundant trophic links (hereafter “redundancy”, see Appendix 1,
Fig. S1, Table S1) from the `r nb_st`
sampling sites with 10+ years of community dynamics (Fig. \@ref(fig:matmet)a-c). We used
ontogenic food diet data at the species level and individual fish body size to
reconstruct trophic species food webs, i.e. at the size class species level (See
Methods). We implemented Bayesian hierarchical models accounting for spatial
variation at both hydrographic river basin and site levels to jointly estimate
the temporal trends in these metrics (see Methods). We found a great
heterogeneity in the temporal trends at the site level (community biomass: from
`r grbmn`% to +`r grbmx`% per decade, species richness: `r grrhn` to
+`r grrhx`%, average trophic level: `r grtln` to +`r grtlx`, connectance:
`r grctn` to +`r grctx`, redundancy: `r grln` to +`r grlx`, Fig.
S2) which was effectively captured by the hierarchical model (conditional R2
ranging from `r r2cond$log_bm_std` for community biomass to
`r r2cond$log_rich_std` for species richness, Table S3).

(ref:sem-cap) **a**, Structural equation models linking the temporal trends of
biodiversity and food web structure. Dashed and solid arrows display
respectively negative and positive the standardized slope coefficients, their
widths being proportional to the absolute values of the coefficients. The
standardized slope coefficients and their associated 95% confidence intervals
obtained by bootstrapping are displayed along the path. $R^2$ displays the marginal
R-squared of the linear models. **b** to **h**, Direct bivariate relationships derived
from the structural equation model. Points display the temporal trends at each
site estimated with a hierarchical Bayesian model, lines display the linear
relationship estimated with the SEM. The distribution of the site temporal
trends are displayed in Fig. S2. All the effects are reported in Table S4.
Average trophic level is weighted by the biomass of the trophic species.
Redundancy is the proportion of redundant links.

@@

(ref:fig-std-cap) **a**, Direct and total effects of community biomass and
species richness trends on food web structure trends. Standardized coefficients
($r_\delta$) derived from the temporal, spatial, and bioenergetic model
Structural Equation Models. Points and lines display the estimated effects and
their confidence intervals at 95%. Values are displayed in Table S4, $R^2$
in Table S5, and the visualization of the bivariate relationships for the
spatial and bioenergetic model SEMs are displayed in Fig. S4. Average trophic
level is weighted by the biomass of the trophic species. Redundancy is the
proportion of redundant links. **b**, Effect of species extinctions on the
structure of the empirical metaweb according to three scenarios. In the
Decreasing and Increasing scenarios, species were ordered by their trophic level
and removed from the metaweb starting respectively from the highest and lowest
trophic level. Species were removed in a random sequence (50 sequences, grey
dots) for the random scenario, the black line displaying the average predicted
by a GAM linear model. For the extinction sequence, the average trophic level
was not weighted by the biomass of species as this information is not available
at the metaweb scale. Similar results were obtained with trophic species (i.e.
node) extinction (Fig. S5).

@@

The coefficients of the Structural Equation Models were computed with
PiecewiseSEM R package [@lefcheck_piecewisesem_2016]. The coefficients were standardized and the
confidence intervals were computed with semEff R package [@murphy_semeff_2021]. Confidence
intervals were obtained by bootstrapping the linear models 1000 times and
computing adjusted bootstrap percentile intervals (BCa), which are adjusted for
the asymmetric distribution of resampling. The marginal and conditional $R^2$
associated with the linear models of the SEMs were computed with the Nagelkerke
method [@lefcheck_piecewisesem_2016, Table S5].

@@

We used a bioenergetic food web model to simulate the relations between
community biomass, species richness and food web structure. This food web model
lifts one of the limitations of our data, that is it includes the biomass of all
species in the food webs from primary producers to top predators. This model
allows simulating complex communities because its parametrization depends
mainly on species body masses. We first generated `r nfw` food webs varying in species
richness (range: `r s_range["min"]`-`r s_range["max"]`)
and connectance (range: `r c_range["min"]`-`r c_range["max"]`) using the niche
model algorithm, which reproduces food web topology typically found in empirical
food webs [@williams_simple_2000]. We set species body masses ($M_i$) using a predator-prey
mass ratio of 100 ($Z$) [@brose_allometric_2006; @brose_predator_2019], such as $M_i =
Z^{(TL_i - 1)}$, $TL_i$ being the trophic level of the species $i$. The
metabolic rates of the species ($x_i$) were then derived from allometric
relationships between metabolism and body mass empirically estimated
[@brose_allometric_2006], such as $x_i = 0.88 \times M_i^{-0.25}$.

@@

# Data, Materials, and Software Availability

The raw data used for the analysis are available on Zenodo:
https://zenodo.org/records/5095656. The raw fish database is public and
available on Zenodo: https://zenodo.org/records/8099409. The manuscript and the
supplementary materials are written in Rmarkdown, i.e. combining code and text,
and are available on github:
https://github.com/alaindanet/fishcom_biomass_dynamics. We further implemented a
code pipeline using the targets R package to ensure that the code, the data, the
figures, the manuscript and the results are up to date.

# Acknowledgments

We are indebted to the OFB for providing the database, and we thank the numerous
people who contributed to the fish records. We thank Dr Katherine Assersohn for
her constructive feedback on the study. AD, LEM, MM, CF, and EL have received
support from by the ANR project ECOSTAB (ANR-17-CE32-0002) awarded to EL. AD and
APB have received support from the Natural Environment Research Council under
grant agreement no. NE/T003502/1 awarded to APB.
